---
title: "Gun Violence in the USA"
output: html_document
date: "2023-05-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/chanks/workspace/gun-violence/gun_violence_archive/mass_shootings_by_year')
```


Summarizing mass shooting data form the gun violence archive, 2014  - 2022: 

```{r}
library(tidyverse)

#loading in the data: 
y2014 = read_csv('gva_2014.csv')
y2015 = read_csv('gva_2015.csv')
y2016 = read_csv('gva_2016.csv')
y2017 = read_csv('gva_2017.csv')
y2018 = read_csv('gva_2018.csv')
y2019 = read_csv('gva_2019.csv')
y2020 = read_csv('gva_2020.csv')
y2021 = read_csv('gva_2021.csv')
y2022 = read_csv('gva_2022.csv')


#creating list of dataframes and years 2014-2022 to iterate through:
my_data = list(y2014,y2015,y2016,y2017,y2018,y2019,y2020,y2021,y2022)
years = seq(2014,2022,1)

for(i in 1:length(my_data)){
 my_data[[i]] = my_data[[i]] %>% mutate(Year = years[i]) #adding a year column for each dataset 
 my_data[[i]] = my_data[[i]] %>% select(-Operations) #removing operations column because value is NA 
}

ms = do.call(bind_rows, my_data) #combining all datasets together
```

Getting summary statistics of mass shootings between 2014 and 2022
```{r}
#number of mass shootings by state 
ms %>% group_by(State) %>% count() %>% arrange(desc(n)) #Illinois is top likely due to gun related violence in Chicaogo. 

ms %>% group_by(Year) %>% count() %>% 
    ggplot(aes(x = Year, y = n)) + 
    geom_point() + geom_line() + 
    labs(title = "The Number of Mass Shootings in the United States has Doubled in 8 Years", 
         subtitle = "An incident is considered a mass shooting if 4 or more people are shot, exluding the shooter",
         x = "Year", 
         y = " # of Mass Shootings", 
         caption = "Source: Gun Violence Archive") + 
    theme_bw()

#is the pandemic to blame for this increase in mass shootings ?
```
Top 10 states with the most mass shootings in the past 8 years
```{r}
ms = ms %>% mutate(State = factor(State)) #changing State to factor to use fct_reorder() 

ms %>% group_by(State) %>% 
  count() %>%
    arrange (desc(n)) %>%
      head(n = 10) %>%
        ggplot(aes(x = fct_reorder(State, n), y = n)) + geom_col() + 
  theme_bw() + 
  labs(title = "Top 10 States with Most Mass Shootings, 2014 - 2018", 
       y = "# of Mass Shootings", 
       x = "State")

#it makes sense that most populous states have most mass shootings. Need to calulate per-capita
#this means I need the state population per year 

```

Getting state population estimates in 2019 from tidycensus 
```{r}
library(tidycensus)

state_pops = get_estimates(geography = "state", 
                           dataset = "pep",
                           variables = 'POP',
                           year = 2019)

state_pops = state_pops %>% rename(State = NAME, pop = value ) %>% select(State, pop)


```

Calculating mass shootings per capita with 2019 state population data
```{r}
ms_2019 = ms %>% filter(Year == 2019) 

ms_2019_by_state = ms_2019 %>% group_by(State) %>% count() %>% rename(num_ms = n)

ms_2019_by_state = ms_2019_by_state %>% left_join(state_pops, by = "State") 
# only 40 states - not every state had a mass shooting in 2019. 

ms_per_cap = ms_2019_by_state %>% mutate(ms_per_cap = num_ms / pop) %>% arrange(desc(ms_per_cap))

```
Top 10 state of mass shootings per capita: 

```{r}
ms_per_cap %>% head(n = 10)
```

Next, we can look at the gun laws in these countries to see if there is a correlation between the high ms states and their lack of gun control. 

```{r}
gun_laws = da37363.0001

gun_laws %>% select(STATE, LAWTOTAL) %>% ggplot(aes(x = fct_reorder(STATE, LAWTOTAL), y = LAWTOTAL)) + geom_col() + coord_flip()


```

Joining mass shooting data with gun law data to determine if correlation exists in 2019 : 

```{r}
gun_laws = gun_laws %>% rename(State = STATE)

gun_laws_2019 = gun_laws %>% filter(YEAR == 2019) %>% group_by(YEAR, State) %>% select(State, LAWTOTAL) 

ms_per_cap = ms_per_cap %>% mutate(State = factor(State))
  
ms_per_cap %>% left_join(gun_laws_2019 %>% select(State, LAWTOTAL, YEAR), by = "State")

ms_per_cap %>% arrange(State)
gun_laws_2019
```

